msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design

### 2025-03-21 paper/manuscript.qmd
Updated manuscript upon peer revision

bibliography: references.bib
format:
  pdf:
    number-sections: true
    number-depth: 3
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
    keep-tex: true
    include-in-header:
      text:
        \usepackage{amsmath, booktabs, caption, longtable, xcolor}
editor: source
crossref:
  fig-title: Figure     # (default is "Figure")
  fig-prefix: Figure
knitr:
  opts_chunk:
    fig.path: ../figs/
---

@@

In this publication, we present a bespoke data analysis workflow for labelled MS-based proteomics experiments, embedded in our msqrob2 universe in R/Bioconductor. Initially designed for label-free workflows, msqrob2 has demonstrated advantages over other state-of-the-art tools in terms of both performance and its modular, transparent implementation [@Sticker2020]. In this work, we extend these benefits to the analysis of labelled experiments. Similar to MSstatsTMT, our workflows address the complex correlation structure inherent in contemporary labelled MS-based proteomics experiments by mixed models. However, unlike MSstatsTMT and msTrawler, we avoid by default the imputation of missing peptide-spectrum match (PSM) intensities, instead assuming missingness at random after accounting for peptide effects. Nevertheless, users retain the option in msqrob2TMT to utilise the default imputation strategies available in QFeatures or to implement custom functions tailored to their specific needs. Additionally, DEqMS and MSstatsTMT model the data upon summarisation at the protein level and msTrawler immediately models the data at the PSM level, while msqrob2TMT provides the user with the option to use either PSM-level or protein-level workflows. Another notable difference is that msqrob2 offers ridge regression and robust M-estimation, which can further stabilise parameter estimation and thus enhance reliable detection of differential proteins.

We demonstrate that our novel msqrob2TMT workflows outperform state-of-the-art tools by benchmarking them against DEqMS [@Zhu2020], MSstatsTMT [@Huang2020], and msTrawler [@msTrawler2024] using two spike-in and real experimental dataset. Our comparisons also highlight that none of the state-of-the-art tools could provide valid inference across all of these three datasets, as their specific implementation limits the experimental designs for which they can infer DA. Particularly, MSstatsTMT cannot account for multiple covariates e.g., treatment effects together with block effects, age and/or other confounders, while DEqMS and msTrawler cannot address the complex correlation in TMT-experiments with multiple runs and technical replication, which are commonly employed in the proteomics community. Note, that Proteome Discoverer was not included in our benchmark, as we utilised datasets featured in the original MSstatsTMT publication, where the authors had already established the superior performance of MSstatsTMT compared to Proteome Discoverer [@Huang2020].

@@

The spike-in dataset was obtained from MassIVE (RMSV000000265) and has the following design: UPS1 peptides at concentrations of 500, 333, 250, and 62.5 fmol were spiked into 50 µg of SILAC HeLa peptides. This series forms a dilution gradient of 1, 0.667, 0.5, and 0.125 relative to the highest UPS1 peptide concentration (500 fmol). A reference sample was prepared by combining the diluted UPS1 peptide samples with 50 µg of SILAC HeLa peptides. Each dilution and the reference sample were processed in duplicate, yielding a total of ten samples. These samples were subsequently labelled using TMT10-plex reagents and combined for LC-MS/MS analysis, hereafter referred to as a "mixture." This protocol was repeated five times. Each mixture was analysed in technical triplicate, resulting in a total of 15 MS runs. This experimental design simulates a scenario with 10 biological replicates per condition, consisting of 2 replicates per condition within each of the five mixtures.

@@

The msTrawler Multibatch Benchmarking Experiment data has been deposited on ProteomeXchange (PXD036799). However, for reproducibility of the results published in the original paper,
the data available on the google drive of the msTrawler paper (https://console.cloud.google.com/storage/browser/mstrawler_paper) was used. The experiment is a spike-in study which consists of a constant mouse plasma background in which yeast proteins were then spiked with the following dilution ratios 1, 2, 3, 5, 11, 14, 18, 20, 24, 28 and 32. The yeast proteins were harvested from different media; glycerol + ammonium sulfate, galactose + urea, galactose + monosodium glutamate, galactose + ammonium sulfate, glucose + ammonium sulfate, glucose + monosodium glutamate, and glucose + urea. The samples were labelled using the TMTpro 18-plex and pooled in six mixtures, each containing 15 samples. Two of the remaining TMT channels are reference channels, one reference channel is the combination of one batch and the other reference channel consists of all batches. For this dataset the ground truth is know: a true positive is a yeast protein that is returned as significant while a significant mouse protein is a false positive.

@@

The DEqMS workflow [@Zhu2020] starts by $\log_2$-transforming the observed PSM intensities. The PSM intensities are then summarised into protein intensities by adopting the median sweep algorithm to the PSM data for each protein in each run, separately. This approach can be regarded as a one-step median polish. Indeed, the $\log_2$-transformed PSM intensities are first centered by subtracting the corresponding median of the $\log_2$ PSM intensities from their corresponding spectrum. Subsequently, the spectrum-centered PSM intensities in each TMT channel for each protein are aggregated into protein expression values using median summarisation. Finally, the protein summaries are normalised by subtracting the median of all summarised protein expression values within their corresponding channel.

@@

We benchmark our novel workflows against state-of-the-art tools by examining the number of reported proteins that are true positive (TP) (i.e., UPS proteins in the spike-in study, and yeast proteins in the Multibatch Benchmark study), and false positive (FP) (i.e., HeLa proteins in the spike-in study, and mouse proteins in the Multibatch Benchmark study). Additionally, we construct true positive rate (TPR)–false discovery proportion (FDP) plots. TPR represents the fraction of truly DA proteins reported by the method, calculated as TPR=TP/(TP+FN), with FN false negatives (i.e., UPS proteins in the spike-in study, and yeast proteins in the Multibatch Benchmark study that were not flagged as DA). FDP denotes the proportion of false positives among all proteins flagged as DA, calculated as FDP = FP/(TP + FP). We also highlight the actually observed FDP at a 5% FDR threshold, which is expected to approximate 5%.

@@

These findings are further corroborated by the results at the 5% FDR level, shown in @fig-WorkflowComparisons C. All msqrob2 workflows demonstrate high sensitivity and low FDP. Specifically, they recover between `r rangeMsqrobTP` spike-in UPS proteins as DA (true positives,TP) while only reporting between `r rangeMsqrobFP` false positives (FP) for HeLa proteins. As a result, their FDP ranges between `r rangeMsqrobFDP`%, suggesting appropriate FDR control at the 5% level. The default msTrawler workflow, however, was only able to recover `r msTrawlerTP` TP and reported `r msTrawlerFP` FP, leading to an FDP of `r msTrawlerFDP`%. With our refactored import function this improved to `r msTrawlerTP_fixed` TP, `r msTrawlerFP_fixed` FP and an FDP of `r msTrawlerFDP_fixed`%. The summarisation-based workflows DEqMS and MSstatsTMT, reported `r deqmsTP` and  `r msStatsTmtTP`  TP, respectively, with `r deqmsFP` and `r msStatsTmtFP` FP, resulting in FDPs of `r deqmsFDP`% and `r msStatsTmtFDP`%, respectively, suggesting improper FDR control by DEqMS.

@@

```{r, echo  =FALSE, warning = FALSE, message  =FALSE, out.width="100%"}
#| label: fig-fold_change_spike
#| fig-cap: "Boxplots showing the log2 FC distributions for the spike-in proteins, focusing on two comparisons with the highest and lowest difference in spike-in concentrations. Fold changes are estimated by DEqMS, msqrob2TMT, MSstatsTMT, and msTrawler workflows. The grey dotted line is the true log2 FC for the comparison. Panel A: non spike-in proteins (yeast); Panel B: spike-in proteins."
#| fig-pos: 'H'
knitr::include_graphics("../figs/figure3.png", error = FALSE)
```

>>>>

bibliography: references.bib
format:
  pdf:
    number-sections: true
    number-depth: 3
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
    keep-tex: true
    include-in-header:
      text:
        \usepackage{amsmath, booktabs, caption, longtable, xcolor, ulem}
editor: source
crossref:
  fig-title: Figure     # (default is "Figure")
  fig-prefix: Figure
knitr:
  opts_chunk:
    fig.path: ../figs/
---

@@

In this publication, we present a bespoke data analysis workflow for labelled MS-based proteomics experiments, embedded in our msqrob2 universe in R/Bioconductor. Initially designed for label-free workflows, msqrob2 has demonstrated advantages over other state-of-the-art tools in terms of both performance and its modular, transparent implementation [@Sticker2020]. In this work, we extend these benefits to the analysis of labelled experiments. Similar to MSstatsTMT, our workflows address the complex correlation structure inherent in contemporary labelled MS-based proteomics experiments by mixed models. However, unlike MSstatsTMT and msTrawler [@msTrawler2024], we avoid by default the imputation of missing peptide-spectrum match (PSM) intensities, instead assuming missingness at random after accounting for peptide effects. Nevertheless, users retain the option in msqrob2TMT to utilise the default imputation strategies available in QFeatures or to implement custom functions tailored to their specific needs. Additionally, DEqMS [@Zhu2020] and MSstatsTMT model the data upon summarisation at the protein level and msTrawler immediately models the data at the PSM level, while msqrob2TMT provides the user with the option to use either PSM-level or protein-level workflows. Another notable difference is that msqrob2 offers ridge regression and robust M-estimation, which can further stabilise parameter estimation and thus enhance reliable detection of differential proteins.

We demonstrate that our novel msqrob2TMT workflows outperform state-of-the-art tools by benchmarking them against DEqMS, MSstatsTMT, and msTrawler using two spike-in and real experimental dataset. Our comparisons also highlight that none of the state-of-the-art tools could provide valid inference across all of these three datasets, as their specific implementation limits the experimental designs for which they can infer DA. Particularly, MSstatsTMT cannot account for multiple covariates e.g., treatment effects together with block effects, age and/or other confounders, while DEqMS and msTrawler cannot address the complex correlation in TMT-experiments with multiple runs and technical replication, which are commonly employed in the proteomics community. Note, that Proteome Discoverer was not included in our benchmark, as we utilised datasets featured in the original MSstatsTMT publication, where the authors had already established the superior performance of MSstatsTMT compared to Proteome Discoverer [@Huang2020].

@@

The spike-in dataset was obtained from MassIVE (RMSV000000265) and has the following design: \sout{UPS1 peptides at concentrations of 500, 333, 250,  and 62.5 fmol}\textcolor{red}{500, 333, 250, and 62.5 fmol of UPS1 peptides} were spiked into 50 µg of SILAC HeLa peptides. This series forms a dilution gradient of 1, 0.667, 0.5, and 0.125 relative to the highest UPS1 peptide \sout{concentration}\textcolor{red}{amount} (500 fmol). A reference sample was prepared by combining the diluted UPS1 peptide samples \textcolor{red}{(286.5 fmol)} with 50 µg of SILAC HeLa peptides. Each dilution and the reference sample were processed in duplicate, yielding a total of ten samples. These samples were subsequently labelled using TMT10-plex reagents and combined for LC-MS/MS analysis, hereafter referred to as a "mixture." This protocol was repeated five times. Each mixture was analysed in technical triplicate, resulting in a total of 15 MS runs. This experimental design simulates a scenario with 10 biological replicates per condition, consisting of 2 replicates per condition within each of the five mixtures. \textcolor{red}{The MS data were searched by the authors using Proteome Discoverer 2.2.0.388 (Thermo Fisher Scientific) and Mascot Server 2.6.1 (Matrix Science, London, UK).}

@@

The msTrawler Multibatch Benchmarking Experiment data has been deposited on ProteomeXchange (PXD036799). However, for reproducibility of the results published in the original paper,
the data available on the google drive of the msTrawler paper (https://console.cloud.google.com/storage/browser/mstrawler_paper) was used. The experiment is a spike-in study which consists of a constant mouse plasma background in which yeast proteins were then spiked with the following dilution ratios 1, 2, 3, 5, 11, 14, 18, 20, 24, 28 and 32. The yeast proteins were harvested from different media; glycerol + ammonium sulfate, galactose + urea, galactose + monosodium glutamate, galactose + ammonium sulfate, glucose + ammonium sulfate, glucose + monosodium glutamate, and glucose + urea. The samples were labelled using the TMTpro 18-plex and pooled in six mixtures, each containing 15 samples. Two of the remaining TMT channels are reference channels, one reference channel is the combination of one batch and the other reference channel consists of all batches. For this dataset the ground truth is known: \textcolor{red}{yeast proteins were spiked in the samples in different amounts, while mouse proteins were added as a constant background. Hence, any yeast protein that is returned as significant is a true positive, and any mouse protein that is returned as significant is a false positive.} \sout{a true positive is a yeast protein that is returned as significant while a significant mouse protein is a false positive}.

@@

The DEqMS workflow [@Zhu2020] starts by $\log_2$-transforming the observed PSM intensities. The PSM intensities are then summarised into protein intensities by adopting the median sweep algorithm to the PSM data for each protein in each run, separately. This approach can be regarded as a one-step median polish \textcolor{red}{that corrects for the PSM-specific effect, that is the effect shared across all TMT intensities within a PSM. More specifically, the median $\log_2$ intensities across TMT are subtracted within each PSM.} \sout{. Indeed, the $\log_2$-transformed PSM intensities are first centered by subtracting the corresponding median of the $\log_2$ PSM intensities from their corresponding spectrum the corresponding PSM.} Subsequently, the spectrum-centered PSM intensities in each TMT channel for each protein are aggregated into protein expression values using median summarisation. Finally, the protein summaries are normalised by subtracting the median of all summarised protein expression values within their corresponding channel.

@@

We benchmark our novel workflows against state-of-the-art tools by examining the number of reported proteins that are true positive (TP) (i.e., UPS proteins in the spike-in study, and yeast proteins in the Multibatch Benchmark study), and false positive (FP) (i.e., HeLa proteins in the spike-in study, and mouse proteins in the Multibatch Benchmark study). Additionally, we construct true positive rate (TPR)–false discovery proportion (FDP) plots. TPR represents the fraction of truly DA proteins reported by the method, calculated as TPR=TP/(TP+FN), with FN false negatives (i.e., UPS proteins in the spike-in study, and yeast proteins in the Multibatch Benchmark study that were not flagged as DA). FDP denotes the proportion of false positives among all proteins flagged as DA, calculated as FDP = FP/(TP + FP). \sout{We also highlight the actually observed FDP at a 5% FDR threshold, which is expected to approximate 5%.} \textcolor{red}{We also highlight the observed FDP at a 5% FDR threshold, computed using the Benjamini-Hochberg (BH) FDR method. Since the FDR represent the expected FDP, that is the average observed FDP if the experiment were to be repeated an infinite number of times, we expect that a method that correctly controls a 5% FDR will lead to an associated observed FDP that is close to 5%.}

@@

These findings are further corroborated by the results at the 5% FDR level, shown in @fig-WorkflowComparisons C. All msqrob2 workflows demonstrate high sensitivity and low FDP. Specifically, they recover \textcolor{red}{across all 6 pairwise comparisons} between `r rangeMsqrobTP` spike-in UPS proteins as DA (true positives,TP) while only reporting between `r rangeMsqrobFP` false positives (FP) for HeLa proteins. As a result, their FDP ranges between `r rangeMsqrobFDP`%, suggesting appropriate FDR control at the 5% level. The default msTrawler workflow, however, was only able to recover `r msTrawlerTP` TP and reported `r msTrawlerFP` FP, leading to an FDP of `r msTrawlerFDP`%. With our refactored import function this improved to `r msTrawlerTP_fixed` TP, `r msTrawlerFP_fixed` FP and an FDP of `r msTrawlerFDP_fixed`%. The summarisation-based workflows DEqMS and MSstatsTMT, reported `r deqmsTP` and  `r msStatsTmtTP`  TP, respectively, with `r deqmsFP` and `r msStatsTmtFP` FP, resulting in FDPs of `r deqmsFDP`% and `r msStatsTmtFDP`%, respectively, suggesting improper FDR control by DEqMS.

@@

```{r, echo  =FALSE, warning = FALSE, message  =FALSE, out.width="100%"}
#| label: fig-fold_change_spike
#| fig-cap: "Boxplots showing the log2 FC distributions for the spike-in and non-spike-in proteins, focusing on two comparisons with the highest and lowest difference in spike-in concentrations. Fold changes are estimated by DEqMS, msqrob2TMT, MSstatsTMT, and msTrawler workflows. The grey dotted line is the true log2 FC for the comparison. Panel A: non spike-in proteins (HeLa); Panel B: spike-in proteins (UPS)."
#| fig-pos: 'H'
knitr::include_graphics("../figs/figure3.png", error = FALSE)
```
